---
title: "Detailed Technical Review: Wilson et al. (2022)"
subtitle: "Synthetic Spike-in Controls for cfMeDIP-seq"
author: "claude.ai Opus 4.5 with prompts from [Francois Collin](https://www.linkedin.com/in/francoisz/)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  # for use of .tabset
  bookdown::html_document2:
  #bookdown::html_book:
  #html_document:
    #code_folding: show
    code_download: true
    toc: true
    toc_depth: 3
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [Ref/_spike_in.bib]
csl: ../../_csl/cell-numeric.csl
link-citations: true
---


```{css sidenote, echo = FALSE}

.main-container {
    margin-left: 250px;
}
.sidenote, .marginnote { 
  float: right;
  clear: right;
  margin-right: -40%;
  width: 37%;         # best between 50% and 60%
  margin-top: 0;
  margin-bottom: 0;
  font-size: 1.1rem;
  line-height: 1.3;
  vertical-align: baseline;
  position: relative;
  }
```


<style>
@import url('https://fonts.googleapis.com/css?family=Raleway');
@import url('https://fonts.googleapis.com/css?family=Oxygen');
@import url('https://fonts.googleapis.com/css?family=Raleway:bold');
@import url('https://fonts.googleapis.com/css?family=Oxygen:bold');

.main-container {
  max-width: 1400px !important;
}

body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1400px; }

caption {
  font-size: 20px;
  caption-side: top;
  text-indent: 30px;
  background-color: lightgrey;
  color: black;
  margin-top: 5px;
}

g-table-intro h4 {
  text-indent: 0px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```

```{r m2b-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r m2b-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 

FN <- "_Wilson_2022_Spike_in_Controls_Technical_Review"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(data.table))
 options(datatable.fread.datatable=F)

 suppressPackageStartupMessages(require(plyr))
 suppressPackageStartupMessages(require(dplyr))
 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


 WRKDIR <- '..'
 if(!file.exists(WRKDIR)) stop("WRKDIR ERROR", WRKDIR)

 # file rmarkdown file management options: cache, figures
 cache_DIR <- file.path(WRKDIR,'Scripts', 'cache/M2B/')
 suppressMessages(dir.create(cache_DIR, recursive=T))
 opts_chunk$set(cache.path=cache_DIR)

 figures_DIR <- file.path(WRKDIR,'Scripts', 'figures/M2B/')
 suppressMessages(dir.create(figures_DIR, recursive=T))
 opts_chunk$set(fig.path=figures_DIR)

 #tables_DIR <- file.path(WRKDIR,'Scripts', 'tables/M2B/')
 #suppressMessages(dir.create(table_DIR, recursive=T))
 #opts_chunk$set(fig.path=table_DIR)
 
 # need a local copy of help_DIR
 #help_DIR <- file.path(WRKDIR,'Scripts', 'help_files')
 help_DIR <- file.path('.', 'help_files')
 suppressMessages(dir.create(help_DIR, recursive=T))
 
 temp_DIR <- file.path(WRKDIR,'Scripts', 'temp_files')
 suppressMessages(dir.create(temp_DIR, recursive=T))

```
<!-- ######################################################################## -->

*** 

```{r m2b-utilityFns, echo=FALSE, eval=FALSE}
 # Here we define some utility functions
source('r/utilityFns.r')

```

<!-- ######################################################################## -->


```{r m2b-input-params-all, echo = FALSE, results = "asis", eval=F, echo=F}
 # print out original input params
data.frame(
param_name=names(params),
param_value=unlist(params),
param_class=sapply(params, class), row.names=NULL) %>%
  knitr::kable(caption="Input Parameters")
```



<style>
    hr {
        border: none;
        height: 2px;
        background: none;
        box-shadow: 0 2px 5px rgba(0,0,0,0.5);
    }
</style>




## Publication Details {-}

**Citation:** Wilson, S.L., Shen, S.Y., Harmon, L., Burgener, J.M., Triche, T., Bratman, S.V., and De Carvalho, D.D. (2022). Sensitive and reproducible cell-free methylome quantification with synthetic spike-in controls. *Cell Reports Methods* 2(9):100294.

**DOI:** 10.1016/j.crmeth.2022.100294

**Authors:** De Carvalho lab (Princess Margaret Cancer Centre), with computational contributions from Michael Hoffman and Tim Triche Jr.

**Significance:** This paper represents a major methodological advancement for cfMeDIP-seq, addressing critical normalization and quantification challenges that had limited the technique's clinical utility.

---

# Background and Motivation

## The Problem

 he authors identified three fundamental limitations in existing cfMeDIP-seq methodology:

1. **Lack of Absolute Quantification**: Prior cfMeDIP-seq studies reported arbitrary read counts rather than absolute amounts of methylated DNA, making cross-sample and cross-experiment comparisons unreliable.

2. **Unaccounted Technical Biases**: DNA fragment properties (length, G+C content, CpG fraction) systematically affect enrichment efficiency, but no standard method existed to correct for these biases.

3. **Batch Effects**: Different laboratories, reagent lots, and processing variations introduced substantial technical variance that confounded biological signals.

## Prior Spike-in Approaches

The original cfMeDIP-seq protocol used **Arabidopsis thaliana DNA** as spike-in controls to assess immunoprecipitation efficiency, but this approach had critical limitations:

- Could not correct for fragment-specific properties
- Provided only global assessment of enrichment
- Did not enable absolute quantification

Other methylation methods used input DNA sequencing (sequencing a portion before enrichment), but this also could not adjust for individual fragment properties affecting enrichment.

---

# Spike-in Design Methodology

## Design Parameters

The authors created a **factorial design** covering the observed range of cfDNA fragment properties:

| Parameter | Values | Rationale |
|-----------|--------|-----------|
| **Fragment Length** | 80 bp, 160 bp, 320 bp | Spanning typical cfDNA size distribution |
| **G+C Content** | 35%, 50%, 65% | Covering low, medium, high GC regions |
| **CpG Fraction** | 1/80 bp, 1/40 bp, 1/20 bp | From CpG-poor to CpG-rich |

This design generates **27 combinations** (3 × 3 × 3 = 27).

## Key Design Features

**Dual Sequences for Each Condition:**

- For each of the 27 combinations, two distinct sequences were designed: one for methylation, one left unmethylated
- Target: **54 total spike-in fragments** (achieved 52 due to synthesis failures)

**Critical Design Constraints:**

1. **No Human Genome Alignment**: All sequences verified with BLASTN against GRCh38/hg38 to ensure no alignment (E-value threshold, bit score <50)

2. **No Secondary Structure**: Checked with UNAFold (80-160 bp) and RNAstructure (320 bp) to avoid synthesis problems and systematic enrichment biases

3. **Markov Model Generation**: Used GenRGenS to generate sequences matching target parameters via first-order Markov models

**Alternative Sequences:**

- Three 320 bp combinations received alternative sequence sets (different sequences, same properties) to assess reproducibility
- These "technical replicates" enabled testing whether sequence content (beyond length/GC/CpG) affected results

## Synthesis and Verification

**Synthesis Approach:**

- 80 bp and 160 bp: Ultramer DNA Oligonucleotides (4 nmol)
- 320 bp: gBlocks Gene Fragments (250 ng)

**Methylation Protocol:**

- PCR amplification of synthetic fragments
- *M.SssI* CpG methyltransferase treatment (37°C for 30 min)
- Verification via restriction digest (HpyCH4IV, HpaII, or AfeI)
- Successful methylation confirmed by single band on 2% agarose gel

**Quantification:**

- Molar amount determined using Qubit dsDNA HS Assay
- Known molecular mass of each fragment used to calculate picomoles

---

# Technical Validation Experiments

## Experiment 1: Spike-ins Alone

**Design:** cfMeDIP-seq performed on 9.99 ng of spike-in DNA pool only (no biological sample)

**Purpose:** Assess inherent biases and performance of spike-ins in isolation

**Key Findings:**

| Metric | Input | Output |
|--------|-------|--------|
| **Methylation Specificity** | 51% methylated, 49% unmethylated | 97% methylated |
| **Non-specific Binding** | N/A | 3% unmethylated |

**Observed Biases (Expected and Confirmed):**

- **Fragment Length**: Enrichment of 160 bp fragments (expected due to size selection for 80-380 bp inserts)
- **G+C Content**: Strong enrichment for high GC fragments
- **CpG Fraction**: Strong enrichment for high CpG density; depletion of CpG-poor fragments (1/80 bp)

**Critical Observation:** Unmethylated fragment signal showed **no association** with length, GC content, or CpG fraction, indicating non-specific binding is random rather than systematic.


## Experiment 2: Optimization with HCT116

**Design:** Spiked varying amounts (0.01 ng, 0.05 ng, 0.1 ng) into 10 ng sheared HCT116 genomic DNA

**Sequencing:** NovaSeq 6000, 60 million paired-end reads per sample

**Key Findings:**

**Optimal Amount: 0.01 ng**

- Used <1% of sequencing reads for spike-ins
- Retained >650,000 spike-in reads for analysis
- Maintained >99% reads for biological sample

**Performance Metrics:**

- Methylation specificity: ≥97%
- Non-specific binding: ≤0.01%
- Consistent bias patterns matching spike-ins-alone experiment

**Reproducibility of Alternative Sequences:**

- Alternative 320 bp fragments with identical properties showed similar read counts
- Largest variation: 50% GC, 1/20 bp CpG fraction
  - Range between alternatives: ≤4% of total reads
  - This represents upper bound on sequence-specific effects

## Experiment 3: Multi-Lab Batch Effect Assessment

**Design:** 5 AML patient plasma samples (10 ng each) + 0.01 ng spike-ins sent to 3 independent labs

**Introduced Variations:**

| Variable | Lab A | Lab B | Lab C |
|----------|-------|-------|-------|
| **Adapter** | xGen Stubby (3-4 bp UMI) | Custom (2 bp UMI) | xGen Stubby (3-4 bp UMI) |
| **Ligation** | 4°C, 16 h | 4°C, 16 h | 20°C, 2 h |
| **Antibody** | Lot RD004 | Lot RD001 | Lot RD004 |
| **Filler DNA** | 50% methylated + 50% unmethylated | 50% methylated + 50% unmethylated | 100% unmethylated |
| **PCR Cycles** | 13-15 (optimized) | 13 | 11 |
| **Sequencer** | NovaSeq (100 bp PE) | NovaSeq (100 bp PE) | NextSeq (75 bp PE) |
| **Read Depth** | 60M | 100M | 85M |

**Critical Finding:** Lab C's use of unmethylated filler (against protocol) resulted in lower methylation specificity, but spike-in normalization successfully corrected this systematic error.

---

# Generalized Linear Model for Absolute Quantification

## Model Specification 

- <span style="color:red"> Should this analysis 
be done on the log scale. </span>

The authors developed a **Gaussian GLM** to predict molar amount (η) from observable features:

```
η = β₀ + β_reads × x_reads + β_len × x_len + β_GC × x_GC + β_CpG × (x_CpG)^(1/3)
```

**Features:**

- **x_reads**: Deduplicated read count (UMI-collapsed)
- **x_len**: Fragment length
- **x_GC**: G+C content (proportion)
- **x_CpG**: CpG fraction (cube root transformed to reduce left skew)

**Output:** Absolute molar amount in picomoles

## Window-Based Analysis

**Genomic Binning:** 300 bp windows across the genome

**Overlap Adjustment:** For fragments partially overlapping windows:

```
η' = (ℓ / ℓ*) × η
```

where:

- ℓ = length of overlap between fragment and window
- ℓ* = 300 bp (window size)
- η' = adjusted molar amount

## Model Performance

**Training/Testing Framework:**

- Used held-out spike-ins to assess prediction accuracy
- Varied training set size from 6 to 25 spike-ins
- 100 random iterations for each training set size

**Results:**

- **Mean absolute error**: ≤0.002 pmol across all test sets
- Performance improved with more training spike-ins
- Larger training sets showed greater variance (due to smaller test sets)
- Model R² = 0.93 for spike-in controls

**Bland-Altman Analysis:** Showed excellent agreement between known and predicted molar amounts, with 95% of predictions within narrow confidence intervals.

---

# Validation Against EPIC Array

## Correlation Analysis

**Comparison:** cfMeDIP-seq molar amounts vs. Illumina EPIC array M-values for HCT116

**Filtering:**

- Removed simple repeats, ENCODE blacklist, low mappability (≤0.5)
- Removed high-variance windows (SD ≥0.05 pmol between replicates)
- **4,446,375 genomic windows** retained after filtering

**Results:**

| CpG Probe Threshold | Molar Amount Correlation | Read Count Correlation |
|---------------------|--------------------------|------------------------|
| ≥3 probes | r = 0.62 | r = 0.61 |
| ≥5 probes | **r = 0.82** | r = 0.80 |
| ≥7 probes | r = 0.87 | r = 0.85 |
| ≥10 probes | r = 0.90 | r = 0.89 |

**Key Insights:**

- Correlation improved with more array probes per window (better representation of true methylation)
- Molar amounts performed slightly better than raw read counts
- Main advantage of molar amounts: **absolute quantification** enables cross-experiment comparisons

**Why M-values?** Authors used M-values rather than β-values due to β-values' high heteroscedasticity (variance dependent on mean).

---

# Batch Effect Mitigation

## Principal Component Analysis

The authors compared four normalization approaches:

1. **Raw read counts** (no normalization)
2. **QSEA normalization** (current standard)
3. **Spike-in molar amounts** (no filtering)
4. **Spike-in + filtering** (recommended approach)

## Key Findings

**Principal Component 1 (PC1):**

| Method | Variance Explained | Associated Variables | Cohen's d Effect Size |
|--------|-------------------|---------------------|---------------------|
| Raw counts | 80% | Batch (non-sig), Filler (non-sig), Adapter (non-sig) | Small |
| QSEA | 81% | **Filler (significant)**, Batch, Adapter | Introduced artifact |
| Spike-in only | 78% | Batch (non-sig) | Small |
| Spike-in + filter | 83% | **Sample (biological)**, **Sex (biological)** | Biological signals dominate |

**Critical Observation:** QSEA normalization **introduced spurious correlation** with filler DNA type, while spike-in normalization removed this technical artifact.

**Principal Component 2/3 (PC2/PC3):**

- Raw data: PC3 (5% variance) significantly associated with batch
- Spike-in + filtering: PC2 (6% variance) associated with batch and filler, but **not confounded with biological variables**

## Variance Reduction

**Batch-Associated Variance:**

- Spike-in controls reduced batch-associated variance to **≤1% of total variance**
- This represents a dramatic improvement in technical reproducibility

**Sample-Sample Correlations:**

- Compared technical replicates (same sample, different labs)
- **10/10 pairings**: Molar amounts showed higher correlation than raw read counts
- Spike-ins improved agreement between labs for identical samples

## Repetitive Element Findings

**PC2 Top Windows (Spike-in + Filtering):**

- Examined top 10% of windows driving PC2 variance (N=258,254)
- **71% mapped to RepeatMasker repetitive elements**
- Of those, **43% were Alu elements**
- Suggests batch effects particularly affect repetitive regions
- These may require special handling in downstream analysis

---

# Quality Control and Filtering

## Methylation Specificity

**Calculation:**
```
Methylation Specificity = (Methylated Reads) / (Total Spike-in Reads)
```

**Recommendation:** Samples with methylation specificity <0.98 should be flagged or removed (indicates poor cfMeDIP performance).

**Implementation:** The spiky package includes `methylation_specificity()` function.

## Recommended Filtering Pipeline

**Remove:**

1. **Simple repeats** (UCSC simple repeats track) - 1,070,387 regions
2. **ENCODE blacklist** - 239,461 regions
3. **Low mappability** (Umap k100 multi-read score ≤0.5) - 549,876 regions
4. **High variance** (SD of molar amount ≥0.05 pmol between replicates) - 906 regions

**Result:** Retained 4,446,375 of original ~5.3M windows (84%)

## High Molar Amount Regions

**Investigation of Outliers:**

- Identified 11 windows with molar amount ≥0.25 pmol
- **4 windows** (≥0.70 pmol): No repetitive elements, located in:
  - *MIER2* (2 windows) - tumor suppressor in colorectal cancer
  - *HCN2*, *RNF126* - genes with tissue-specific expression
- **7 windows** with mappability <1: Overlap repetitive elements
  - 6/7 overlapped **Alu elements**
  - Likely represent regions with extra copies not in reference genome

**Biological Insight:** Top-ranking non-repetitive regions overlap cancer-relevant genes with altered expression/methylation in HCT116 colon cancer cells.

---

# Computational Tools: The spiky Package

## Package Overview

**Repository:** Available on Bioconductor and GitHub (https://github.com/trichelab/spiky)

**Purpose:** End-to-end processing of cfMeDIP-seq data with spike-in controls

**Language:** R (version ≥3.6 required, ≥4.0 recommended)

## Core Functions

**1. methylation_specificity()**

- Calculates methylation specificity from spike-in coverage
- Returns mean and median values
- Flags poor-quality samples

**2. model_glm_pmol()**

- Builds Gaussian GLM on methylated spike-ins only
- Generates batch-specific coefficients
- Creates Bland-Altman diagnostic plot
- Returns trained model object

**3. Prediction Pipeline**

- Applies trained model to biological samples
- Predicts molar amount for each genomic window
- Adjusts for fragment overlap with windows
- Outputs absolute quantification in picomoles

## Workflow Integration

**Input:** Coverage objects from aligned cfMeDIP-seq BAM files

**Process:**

1. Load spike-in coverage data
2. Calculate methylation specificity
3. Train GLM on spike-in data
4. Predict molar amounts for biological samples
5. Apply filtering recommendations
6. Output quantified methylation data

---

# Data Availability

## Published Data

**GEO Accession GSE166259:**

- Raw HCT116 cell line data (public)
- Processed AML patient data (public)

**EGA Accession EGAS00001005069:**

- Raw AML patient data (controlled access)
- Available via University Health Network Genomics Data Access Committee

**Zenodo Deposits:**

- Code for reproducing all analyses: https://doi.org/10.5281/zenodo.4683791
- spiky package: https://doi.org/10.18129/B9.bioc.spiky

## Spike-in Sequences

**Table S1:** Complete sequences of all 52 spike-in fragments with:

- Sequence ID
- Fragment length
- G+C content
- CpG fraction
- Methylation status
- Optimal annealing temperature
- Restriction enzymes for methylation verification

**Commercial Availability:** All sequences can be ordered from commercial providers (Integrated DNA Technologies)

---

# Patent and Commercial Considerations

**Patent Application:** PCT/CA2020/051507 filed by Wilson, Shen, Triche, De Carvalho, and Hoffman

**License:** Licensed to **Adela, Inc.**

**Conflict of Interest:**

- Shen, Bratman, and De Carvalho: Leadership roles and equity in Adela
- Multiple authors: Inventors on related cfDNA methylation patents licensed to Adela

**Implication:** The spike-in control technology is protected IP but apparently available for research use through the open-source spiky package.

---

# Key Innovations and Contributions

## Methodological Advances

1. **First synthetic spike-ins for enrichment-based methylation methods**
   - Prior art existed for RNA-seq, ChIP-seq, bisulfite methods
   - No prior spike-ins designed for MeDIP-seq or cfMeDIP-seq

2. **Comprehensive factorial design**
   - Systematically covers parameter space of fragment properties
   - Enables correction for length, GC, and CpG biases simultaneously

3. **Absolute quantification capability**
   - Converts arbitrary read counts to picomoles
   - Enables true quantitative comparisons across experiments

4. **Batch effect mitigation**
   - Reduces batch-associated variance to <1% of total
   - Preserves biological signals while removing technical artifacts

## Quality Control Metrics

1. **Methylation specificity**: Sample-level QC
2. **Unmethylated spike-in signal**: Measures non-specific binding
3. **Alternative sequence reproducibility**: Internal consistency check
4. **Model fit diagnostics**: Bland-Altman plots for QC

---

# Limitations and Future Directions

## Acknowledged Limitations

**1. Limited Parameter Coverage**

- Only 3 values for each parameter (length, GC, CpG)
- Future designs with more granular coverage could improve modeling

**2. Binary Methylation Assessment**

- Spike-ins are either fully methylated or unmethylated
- Cannot distinguish methylation status of individual biological fragments post-enrichment
- Know ~97% are methylated, but not which specific fragments

**3. Partial CpG Coverage**
- 320 bp fragments with 65% GC failed synthesis due to secondary structure
- Missing one parameter combination from full factorial design

## Unstated Limitations and Considerations

**1. Platform Specificity**

- Designed specifically for Illumina paired-end sequencing
- Adapter sequences are platform-specific
- May require redesign for other sequencing platforms (PacBio, Oxford Nanopore)

**2. Sample Type Assumptions**

- Validation performed on plasma cfDNA
- Performance with other cfDNA sources (CSF, urine, saliva) not validated

**3. Cancer Type Generalization**

- Validation used AML (blood cancer with high cfDNA levels)
- Performance with low cfDNA cancers (early-stage solid tumors) requires validation

**4. Input Amount Optimization**

- 0.01 ng spike-in optimized for 10 ng biological sample
- May need adjustment for very low input samples (<10 ng)

**5. Computational Requirements**

- GLM requires sufficient spike-in reads (>650,000)
- May be challenging with low-depth sequencing

---

# Impact on the Field

## Immediate Applications

**1. Clinical Assay Development**

- Enables FDA/regulatory submissions with absolute quantification
- Provides standardized QC metrics
- Facilitates multi-center studies with consistent normalization

**2. Research Studies**

- Meta-analyses now possible with common quantification units
- Improved statistical power through variance reduction
- Better detection of small biological effects

**3. Technology Development**

- Provides foundation for assay optimization
- Enables systematic comparison of protocol variations
- Quantifies impact of reagent changes

## Adoption by Adela

This technology has been incorporated into Adela's commercial MRD and MCED platforms, as evidenced by:

- Patent licensing to Adela
- Co-authors' involvement with Adela
- Integration with MEDIPIPE pipeline (Zeng et al. 2023)

## Broader Implications

**For Enrichment-Based Methods:**

- Design principles applicable to other MeDIP-seq variants
- Could be adapted for histone modification ChIP-seq
- Template for developing spike-ins for other enrichment assays

**For cfDNA Field:**

- Demonstrates importance of technical standardization
- Provides model for rigorous validation studies
- Raises bar for commercial assay development

---

# Technical Strengths

## Experimental Design

1. **Systematic approach**: Factorial design ensures comprehensive coverage
2. **Multiple validation experiments**: Spike-ins alone → HCT116 optimization → Multi-lab batch testing
3. **Appropriate controls**: Unmethylated spike-ins assess non-specific binding
4. **Biological validation**: Correlation with orthogonal platform (EPIC array)

## Statistical Rigor

1. **Held-out validation**: Model tested on unseen spike-ins
2. **Multiple iterations**: 100 random splits for robustness
3. **Effect size reporting**: Cohen's d for PCA associations
4. **Multiple testing correction**: Holm-Bonferroni adjustment
5. **Diagnostic plots**: Bland-Altman for model assessment

## Open Science

1. **Complete data deposition**: Raw and processed data publicly available
2. **Open-source software**: spiky package on Bioconductor
3. **Detailed methods**: Nature Protocols-level protocol description
4. **Reproducible code**: All analysis code on Zenodo/GitHub

---

# Connection to cfMeDIP-seq Research Documentation

## Relevance to Comprehensive cfMeDIP-seq Documentation

**Methodological Integration:**

- This paper provides the **quantification layer** missing from standard documentation
- Complements MEDIPIPE pipeline (which incorporates spike-in normalization in later versions)
- Addresses variance sources in protocol optimization

**Documentation Updates:**

1. Add spike-in design as a protocol step
2. Include GLM-based quantification in computational pipeline
3. Incorporate methylation specificity as QC metric
4. Add batch effect mitigation section

## Relevance to MCED Comparison (Adela vs GRAIL)

**For Adela Platform:**

- This spike-in technology is **foundational** to Adela's cfMeDIP-seq assay
- Explains how Adela achieves absolute quantification
- Provides mechanistic basis for batch effect resistance
- Supports claims of high specificity (0.01% non-specific binding)

**For GRAIL Comparison:**

- GRAIL uses targeted bisulfite sequencing (no spike-ins mentioned in literature)
- Adela's spike-in approach enables:
  - Absolute quantification (picomoles vs. arbitrary counts)
  - Genome-wide bias correction
  - Superior batch effect mitigation (<1% variance)
- This represents a technical differentiator in platform comparison

## For Enrichment Methods Section

**Should Be Included:**

- **Section 2.3.6**: "Spike-in Controls for Absolute Quantification"
  - Design principles
  - GLM modeling approach
  - Quality control metrics
  - Batch effect mitigation

**Integration with Existing Content:**

- Complements MEDIPIPE coverage (MEDIPIPE incorporates spike-in support)
- Provides mathematical basis for normalization approaches
- Explains variance reduction mechanisms

## Key Equations to Document


**1. GLM Quantification Model:**
```
η = β₀ + β_reads × x_reads + β_len × x_len + β_GC × x_GC + β_CpG × (x_CpG)^(1/3)
```

**2. Window Overlap Adjustment:**
```
η' = (ℓ / ℓ*) × η
```

**3. Methylation Specificity:**
```
Specificity = N_methylated / N_total
```

---

# Recommended Citations for Documentation

**Primary Citation:**
Wilson, S.L., Shen, S.Y., Harmon, L., et al. Sensitive and reproducible cell-free methylome quantification with synthetic spike-in controls. *Cell Rep Methods* 2022;2(9):100294. doi:10.1016/j.crmeth.2022.100294

**Related Methods:**
- Shen et al. 2019 (*Nat Protoc*): Original cfMeDIP-seq protocol (doi:10.1038/s41596-019-0202-2)
- Zeng et al. 2023 (*Bioinformatics*): MEDIPIPE with spike-in integration (doi:10.1093/bioinformatics/btad423)

**Software:**
- spiky R package (Bioconductor): https://bioconductor.org/packages/spiky
- GitHub repository: https://github.com/trichelab/spiky

**Data:**
- GEO: GSE166259 (public data)
- EGA: EGAS00001005069 (controlled access)
- Zenodo: 10.5281/zenodo.4683791 (analysis code)

---

# Critical Assessment

## Strengths

1. **Rigorous validation** across multiple experimental systems
2. **Open and reproducible** research practices
3. **Practical impact** through commercial adoption
4. **Elegant solution** to complex normalization problem
5. **Well-documented** methods and software
6. **Strong statistical framework** with held-out validation
7. **Comprehensive QC metrics** for sample quality assessment

## Areas for Further Development

1. **Long-read compatibility**: Current design optimized for short-read sequencing
2. **Ultra-low input**: Performance with <10 ng input not validated
3. **Alternative enrichment methods**: Adaptation to MBD-seq or other approaches
4. **Fully automated QC**: Integration with standard pipelines could be improved
5. **Cost analysis**: Economic impact of spike-in addition not discussed
6. **Clinical validation**: Performance in prospective clinical studies not yet reported

---

# Experimental Timeline and Sample Processing

## Sample Preparation Time Requirements

**Spike-in Preparation (One-time):**

- Synthesis: 2-3 weeks (commercial)
- PCR amplification: 2-3 hours
- Methylation: 1 hour
- Verification: 2-3 hours
- Quantification: 30 minutes

**Per-Experiment Addition:**

- Spike-in addition to samples: <5 minutes
- No additional library prep time

## cfMeDIP-seq Protocol Timeline with Spike-ins

**Day 1:**

- DNA extraction and quantification: 2-3 hours
- Spike-in addition: 5 minutes
- End-repair and A-tailing: 1.5 hours
- Adapter ligation: overnight at 4°C

**Day 2:**

- Bead cleanup: 1 hour
- DNA denaturation: 30 minutes
- Save input (10%): 5 minutes
- Immunoprecipitation: 2 hours
- Washing: 1 hour
- DNA elution: 1 hour
- Library amplification: 2 hours
- Size selection: 1 hour

**Day 3:**

- Quality control: 1-2 hours
- Sequencing submission

**Total Hands-on Time:** ~12-14 hours over 3 days
**Spike-in Overhead:** <5 minutes per sample

---

# Computational Pipeline Requirements

## Software Dependencies

**Core Tools:**

- fastp (v0.11.5+): Adapter trimming, UMI handling
- Bowtie2 (v2.3.5+): Alignment
- UMI-tools (v1.0.0+): Deduplication
- samtools (v0.10.2+): BAM processing
- bedtools (v2.29.2+): Window overlap calculations

**R Packages:**

- spiky (Bioconductor): Spike-in analysis
- stats: GLM fitting
- sesame (v1.8.2+): Array data processing (optional)

**Reference Data:**

- Spike-in sequences (52 fragments)
- Human reference genome (GRCh38/hg38)
- UCSC simple repeats
- ENCODE blacklist
- Umap mappability scores (k100)

## Computational Resources

**Minimum Requirements:**

- CPU: 8 cores
- RAM: 32 GB
- Storage: 100 GB per sample (raw data)
- Processing time: 2-4 hours per sample

**Recommended:**

- CPU: 16+ cores
- RAM: 64 GB
- Storage: SSD for working directory
- HPC cluster for batch processing

---

# Quality Control Thresholds and Decision Rules

## Sample-Level QC

**Methylation Specificity:**

- **Pass:** ≥0.98
- **Warning:** 0.95-0.98
- **Fail:** <0.95

**Spike-in Read Count:**

- **Pass:** ≥500,000 reads
- **Warning:** 250,000-500,000 reads
- **Fail:** <250,000 reads

**Non-specific Binding:**

- **Pass:** ≤0.01
- **Warning:** 0.01-0.03
- **Fail:** >0.03

## Window-Level QC

**Standard Deviation Threshold:**

- <span style="color:red">Remove windows with SD ≥0.05 pmol between replicates
</span>
</span>
- Indicates inconsistent quantification

**Mappability Threshold:**

- Remove windows with Umap k100 score ≤0.5
- Indicates multimapping concerns

**Blacklist Filtering:**

- Remove all ENCODE blacklist regions
- Remove all UCSC simple repeats

## Model Quality Assessment

**GLM Fit Quality:**

- R² ≥0.90 (excellent)
- R² 0.85-0.90 (good)
- R² <0.85 (investigate)

**Mean Absolute Error:**

- ≤0.002 pmol (excellent)
- 0.002-0.005 pmol (acceptable)
- >0.005 pmol (investigate)

---

# Comparison with Alternative Normalization Methods

## Normalization Strategies

| Method | Approach | Advantages | Limitations |
|--------|----------|------------|-------------|
| **Spike-in (Wilson 2022)** | Synthetic controls + GLM | Absolute quantification, batch correction, bias adjustment | Requires spike-in synthesis, additional cost |
| **QSEA** | Computational model | No additional reagents, established method | May introduce artifacts, no absolute quantification |
| **Input Normalization** | Sequencing pre-IP DNA | Simple, low cost | Cannot adjust fragment-specific biases |
| **Total Read Depth** | Library size scaling | Computationally simple | Confounds biological with technical variation |
| **Spike-in (Arabidopsis)** | Plant DNA controls | Simple implementation | Cannot correct fragment properties, no absolute quantification |

## Performance Metrics

**Batch Effect Reduction:**

- Spike-in: <1% of total variance
- QSEA: ~5% of total variance (introduces filler DNA artifact)
- Raw counts: ~5% of total variance

**Correlation with Arrays:**

- Spike-in: r = 0.82 (≥5 probes)
- Raw counts: r = 0.80 (≥5 probes)
- Difference is in absolute quantification capability

---

# Integration with MEDIPIPE Pipeline

## MEDIPIPE Overview

MEDIPIPE (Zeng et al. 2023) is an automated pipeline for cfMeDIP-seq that incorporates spike-in control support.

**Key Features:**
- Automated QC from FASTQ to results
- Integration with spiky package
- Multi-sample processing
- Quality metric reporting

## Workflow Integration Points

**Stage 1: Alignment**

- FASTQ → BAM alignment to spike-in sequences
- Separate alignment to human genome
- UMI-based deduplication

**Stage 2: Spike-in Analysis (spiky)**

- Calculate methylation specificity
- Train GLM model
- Generate QC plots

**Stage 3: Quantification**

- Apply GLM to biological samples
- Window-based molar amount calculation
- Generate normalized count matrices

**Stage 4: Filtering**

- Apply recommended filters
- Output analysis-ready data

---

# Cost-Benefit Analysis

## Additional Costs

**One-Time Costs:**

- Spike-in synthesis: ~$2,000-3,000 for 52 fragments
- Methylation reagents: ~$200
- Validation testing: ~$500-1,000

**Per-Sample Costs:**

- Spike-in DNA: ~$0.10-0.50 per sample (0.01 ng)
- Additional sequencing: Negligible (<1% of reads)
- Computational: Minimal (automated)

**Total Additional Cost:** ~$1-2 per sample after initial investment

## Benefits

**Quantitative:**

- Batch effect reduction: 5% → <1% variance
- Reproducibility: 10/10 technical replicates improved
- Absolute quantification: Enables cross-study comparisons

**Qualitative:**

- Regulatory compliance: Standardized QC metrics
- Multi-center studies: Consistent normalization
- Clinical validation: Reduced technical noise improves signal detection

**Return on Investment:**

- Critical for clinical assay development
- Reduces need for repeat experiments
- Enables meta-analyses across studies

---

# Future Developments and Opportunities

## Technical Improvements

**Expanded Spike-in Libraries:**

- More granular parameter coverage (5-7 values per parameter)
- Additional fragment sizes (40 bp, 60 bp, 200 bp, 400 bp)
- Different methylation patterns (partial methylation)

**Platform Extensions:**

- Long-read sequencing adaptation (PacBio, ONT)
- Single-cell cfMeDIP-seq
- Low-input protocols (<1 ng)

**Computational Advances:**

- Machine learning models beyond GLM
- Real-time QC during sequencing
- Automated troubleshooting recommendations

## Application Extensions

**Other Enrichment Methods:**

- MBD-seq (methyl-CpG binding domain)
- MethylCap-seq (MBD protein capture)
- Hydroxymethylation enrichment (5hmC)

**Histone Modifications:**

- ChIP-seq for histone marks
- CUT&RUN/CUT&Tag applications
- ATAC-seq optimization

**Other Sample Types:**

- Tissue DNA (FFPE, fresh frozen)
- Urine cfDNA
- CSF cfDNA
- Saliva cfDNA

## Clinical Translation

**Regulatory Pathways:**

- FDA submissions with validated QC metrics
- Clinical laboratory certification (CLIA/CAP)
- International standardization (ISO)

**Clinical Studies:**

- Prospective validation in MCED trials
- MRD detection performance
- Tissue-of-origin prediction accuracy

**Commercial Development:**

- Kit-based spike-in controls
- Automated analysis software
- Reference materials for proficiency testing

---

# Summary and Conclusions

## Key Achievements

Wilson et al. (2022) represents a **landmark methodological advance** for cfMeDIP-seq technology. The synthetic spike-in control system solves three critical problems:

1. **Absolute quantification**: Converts arbitrary counts to picomoles
2. **Bias correction**: Adjusts for fragment-specific enrichment efficiency
3. **Batch effect mitigation**: Reduces technical variance to <1% of total

## Technical Excellence

The work demonstrates **exceptional rigor** in:

- **Experimental design**: Factorial approach with systematic validation
- **Statistical methodology**: Held-out testing, effect size reporting, multiple testing correction
- **Open science**: Complete data/code sharing, open-source software
- **Practical implementation**: spiky R package for widespread adoption

## Clinical and Commercial Impact

**Commercial Adoption:**

- Licensed to Adela, Inc. for MCED and MRD applications
- Incorporated into MEDIPIPE analysis pipeline
- Foundation for regulatory submissions

**Research Impact:**

- Enables meta-analyses across studies
- Facilitates multi-center collaborations
- Raises standards for assay development

## Relevance to Current Research

**For cfMeDIP-seq Documentation:**

- Essential for understanding quantification methodology
- Provides framework for QC metrics
- Explains batch effect handling strategies

**For MCED Platform Comparison:**

- Key differentiator between Adela and GRAIL platforms
- Explains Adela's absolute quantification capability
- Supports claims of technical reproducibility

**For Enrichment Methods Analysis:**

- Provides template for normalization approaches
- Demonstrates importance of technical controls
- Offers blueprint for other enrichment assays

---

# References

## Primary Citation

Wilson, S.L., Shen, S.Y., Harmon, L., Burgener, J.M., Triche, T., Bratman, S.V., and De Carvalho, D.D. (2022). Sensitive and reproducible cell-free methylome quantification with synthetic spike-in controls. *Cell Reports Methods* 2(9):100294. doi:10.1016/j.crmeth.2022.100294

PMID: 36160046  
PMCID: PMC9499995

### 26.2 Key Related Publications

**Original cfMeDIP-seq Method:**

- Shen, S.Y., Singhania, R., Fehringer, G., et al. (2018). Sensitive tumour detection and classification using plasma cell-free DNA methylomes. *Nature* 563(7732):579-583. doi:10.1038/s41586-018-0703-0

**cfMeDIP-seq Protocol:**

- Shen, S.Y., Burgener, J.M., Bratman, S.V., and De Carvalho, D.D. (2019). Preparation of cfMeDIP-seq libraries for methylome profiling of plasma cell-free DNA. *Nat Protoc* 14(10):2749-2780. doi:10.1038/s41596-019-0202-2

**MEDIPIPE Pipeline:**

- Zeng, Y., Ye, W., Stutheit-Zhao, E.Y., et al. (2023). MEDIPIPE: an automated and comprehensive pipeline for cfMeDIP-seq data quality control and analysis. *Bioinformatics* 39(7):btad423. doi:10.1093/bioinformatics/btad423

**Alternative Normalization (QSEA):**

- Lienhard, M., Grasse, S., Rolff, J., et al. (2017). QSEA—modelling of genome-wide DNA methylation from sequencing enrichment experiments. *Nucleic Acids Res* 45(6):e44. doi:10.1093/nar/gkw1193

## Software and Data Resources

**Software:**

- spiky R package: https://bioconductor.org/packages/spiky
- GitHub repository: https://github.com/trichelab/spiky
- Analysis code: https://github.com/hoffmangroup/2020spikein

**Data:**

- GEO GSE166259: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE166259
- EGA EGAS00001005069: https://ega-archive.org/studies/EGAS00001005069
- Zenodo: https://doi.org/10.5281/zenodo.4683791

**Patent:**
- PCT/CA2020/051507: Synthetic spike-in controls for cell-free MeDIP sequencing

---

# Glossary of Key Terms

**Absolute Quantification:** Measurement of methylated DNA in defined units (picomoles) rather than arbitrary read counts.

**Batch Effect:** Systematic technical variation introduced by different processing times, reagent lots, laboratories, or operators.

**CpG Fraction:** The number of CpG dinucleotides per unit length of DNA fragment (e.g., 1/80 bp = 1 CpG per 80 base pairs).

**Generalized Linear Model (GLM):** Statistical model that relates response variable (molar amount) to predictor variables (read count, length, GC content, CpG fraction) through a link function.

**M-value:** Log2 ratio of methylated to unmethylated intensities from methylation arrays; preferred over β-values due to homoscedasticity.

**Methylation Specificity:** Proportion of enriched fragments that are methylated; used as quality control metric (target ≥0.98).

**Non-specific Binding:** Capture of unmethylated DNA fragments during immunoprecipitation; measured using unmethylated spike-ins.

**Spike-in Controls:** Synthetic DNA fragments of known sequence, quantity, and methylation status added to samples before processing.

**UMI (Unique Molecular Identifier):** Short random sequences added during library preparation to identify PCR duplicates and enable deduplication.

**Umap Mappability:** Quantitative measure of how uniquely a genomic region can be mapped by sequencing reads of given length.

---

*Document prepared: January 2026*

*Review by: Research Assistant specializing in DNA methylation and liquid biopsy technologies*

*Note: This technical review is intended for research purposes. The spike-in control technology is subject to patent protection (PCT/CA2020/051507) licensed to Adela, Inc.*



<!-- To run
## nohup Rscript -e "knitr::knit2html('_Wilson_2022_Spike_in_Controls_Technical_Review.Rmd')" > _Wilson_2022_Spike_in_Controls_Technical_Review.log  &

## Or
## nohup Rscript -e "rmarkdown::render('_Wilson_2022_Spike_in_Controls_Technical_Review.Rmd')" > _Wilson_2022_Spike_in_Controls_Technical_Review.log  &

-->
_Wilson_2022_Spike_in_Controls_Technical_Review

