---
title: "Adela's cfMeDIP-seq Technology"
#subtitle:
author: "Francois Collin ([CV](https://francoiscollin-cv.netlify.app), [linkedin](https://www.linkedin.com/in/francoisz/))"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  #html_document:
  # for use of .tabset
  rmarkdown::html_document:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 2
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../_bibFiles/_cfMeDIP-seq.bib, ../_bibFiles/_misc.bib, ../_bibFiles/_AI_ML.bib, ../_bibFiles/_RUV.bib, ../_bibFiles/_scRNAseq_norm.bib, ../_bibFiles/_ref_free_genom_inf.bib, ../_bibFiles/_healthy_aging.bib, ../../_bibFiles/_Normalization.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Freedman.bib, ../../_bibFiles/_Yu.bib,  ../../_bibFiles/_RMA.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

<p style="background-color: yellow; padding: 5px;">
  <span style="font-weight: bold;">⚠️ Work in Progress:</span> This content is incomplete.
</p>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```


```{r infex-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r index-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 


FN <- "index"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "index" #- replace by FLOWCELL???

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


```
<!-- ######################################################################## -->

<br/>

<!--
* Donoho on Frictionless Reproducibility
   - Donoho on the Modern History of Stats
-->

# About this page {-}

* In this page we review the literature
which document the development of Adela's cfMeDIP-seq Technology.
The source of all the information contained therein is public.


<br/>

# AI Reports 

* AI reports are produced by claude.ai Opus 4.5 with 
naive prompts from Francois Collin.

* The reporting done is intended to provide the information
needed to understand the roots and development of 
Adela's cfMeDIP-seq Technology.



<br/>

## [Adela vs GRAIL](AI/_Adela_vs_GRAIL_MCED_Comparison.html)

* A fairly comprehensive, high level comparison of the two platforms.

* This compilation is for background purposes - see the documentation below 
for comments on technical aspects of some of the studies cited.

<br/>

## [cfMeDIP-seq: Protocol, Pipeline and Classifiers](AI/_cfMeDIP-seq_Protocol_Pipeline_Classifiers.html)

<p/>
* The evolution of Adela's cfMeDIP-seq Protocol, Pipeline and Classifiers is 
traced.
* The report is a fairly comprehensive, high level description of the assay
and algorithm along with a description of enhancements that followed.


<br/>

## [Differential Methylation Analysis Methodology](AI/_Differential_Methylation_Analysis_Report.html)

<p/>
* Exhaustive survey of DMA methods used with each  assay type. 
   - Not specific to Adela
* Describe the analyis data in each step
* Survey of benchmarking studies and tools

<br/>

## [Enrichment-Based Differential Methylation Analysis](AI/_Enrichment_Methods_PartI_Section1.html)

<p/>
* For enrichment-Based assays, a look at Bioinformatics Pipelines and 
Statistical Methods for Differential Methylation Analysis: 
   - Common Bioinformatics Pipeline:FASTQ->BAN[ ](AI/_Enrichment_Methods_PartI_Section1.html#common-bioinformatics-pipeline-fastq-to-bam)
   - DMA Methods for Enrichment based assays[ ](AI/_Enrichment_Methods_PartI_Section1.html#dma-methods)
      - many of the methods produce estimates of `absolute beta`, a measure of
methylation level.  These would then be further analyzed to
get DMA results.

<br/>
 

# RF vs elastic net

* RF vs elastic net: does RF edge enet and why?

<p/>
* From personal experience enet should do better with RF performance
dropping substantially in the validation/evaliation set.
* Also from [Statistical learning with big data. A talk by Trevor Hastie](https://
outu.be/0EWJZIC4JxA?si=rGc6_66nODHE3fTP&t=485) (University of Bristol):
   - wide data - many features but few observations - calls for special
methods to avoid overfitting.  The elastic net is one of those methods.
   - tall data - many observations but reasonable number of features -
not at risk of overfitting and can fit richer models with high orders
of interactions.

<span style="color:green"> 
*  Rerun the analyses which fit predictive models and
compare RF and enet resultsA (Halla-aho & Lähdesmäki (BMC Bioinformatics)** [@Hallaaho2022]
.
   - RF is expected to lag in performance because the fitting dataset
is not ruch enough to support the more complex RF models.

* Word on feature selection - it is not needed here.
   - Predicting cancer is not a feature  selection ploblem;
it's a predictive analysis problem.
</span>
 
<br/>

# Variability in  cfMeDIP-seq Data

<p/>
* The key components of the assay - filler DNA and antibody instance - 
are associated with a considerable amount of variability.
   - [docname](./docname.html) provides a detailed look at
the Methylated DNA Immunoprecipitation Stage of the protocol.
   - questions to address:
      - how does the filler DNA operate?  where does the variability
come from?
      - how does anti-5-methylcytosine (5mC) antibody work?
where does the variablity come from?
      - can variability be attributed to its proper source -
anti-5-methylcytosine (5mC) antibody or filler DNA?

<span style="color:green">
*  Rerun the analyses which fit predictive models and
compare RF and enet results.
   - RF is expected to lag in performance because the fitting dataset
is not ruch enough to support the more complex RF models.

* Word on feature selection - it is not needed here.
   - Predicting cancer is not a feature  selection ploblem;
it's a predictive analysis problem.
</span>

<br/>

# References 

<div id="refs"></div>

<br/>


```{r, echo=FALSE}
  knitr::knit_exit()
```

############################################################
# ARCHIVED CODE BELOW
############################################################

# Appendix: Full list of reports

```{r , child=c('_index.Rmd')}
```


## Full List of AI Reports


```{r , child=c('_index_AI.Rmd')}
```

<br/>

### Study by Study Execution Outline  {-}

#### **Shen et al. (2018) [@Shen:2018aa]** {-}

* Was the Differentially Methylated Regions (DMRs) status between cancer and controls
used for selecting features?

* CV is to determine the hyper-parameters

```
Training:

* 10-fold cross-validation for hyper-parameter tuning
* Nested cross-validation for performance estimation
* Class imbalance handled by:
   - Weighted loss function
   - Or downsampling of majority class

```
 
* class imbalance is not an issue when comparing algorithms based on the same sample set.

<p><p/>
* auROC is not a good summary
   - spec at fixed sens would be better
   - but both would order algorithms similarly.



<br/>

### Controls, Controls, Controls 

<p><p/>
* Spike-In Controls - Wilson et al. (2022) [@Wilson:2022aa]
   - How are the Synthetic Spike-In Controls used to reduce unwanted variability?


   - Spike-ins may not be affected by some important sources of variability,
but they can only contribute adjustments for variability it is subjected to.
   - What else could be done with these?  
   - What other controls could be exploited? 
      - endogenous controls
      - sample controls

<br/>

### Before any modeling, fitting, etc. 

<p><p/>
* We need to examine the data to get a sense of what the problem is:
   - are there obvious quality issues?
   - are there less obvious quality issues?  
   - visually, how hard of a classification problem  do we have?
      - are cancer types distinct?
      - are there outlier points?  Is the outlier due to technical reasons?
 
<br/>

<!--
#### Word on modeling

* It is the error structure in the dataset that dictates the properties 
and validity of the computed statistics 

* Combining data points works well statistically if they are independent with
common mean and variance (+/-)

* statistical modeling is essentially the problem of finding                    contrasts among variables or transformations of variables such that
the contrasts are independent with common mean and sd.

-->


## Full List of AI Reports


```{r , child=c('_index_AI.Rmd')}
```

<br/>

### Study by Study Execution Outline  {-}

#### **Shen et al. (2018) [@Shen:2018aa]** {-}

* Was the Differentially Methylated Regions (DMRs) status between cancer and controls
used for selecting features?

* CV is to determine the hyper-parameters

```
Training:

* 10-fold cross-validation for hyper-parameter tuning
* Nested cross-validation for performance estimation
* Class imbalance handled by:
   - Weighted loss function
   - Or downsampling of majority class

```
 
* class imbalance is not an issue when comparing algorithms based on the same sample set.

<p><p/>
* auROC is not a good summary
   - spec at fixed sens would be better
   - but both would order algorithms similarly.



<br/>

### Controls, Controls, Controls 

<p><p/>
* Spike-In Controls - Wilson et al. (2022) [@Wilson:2022aa]
   - How are the Synthetic Spike-In Controls used to reduce unwanted variability?


   - Spike-ins may not be affected by some important sources of variability,
but they can only contribute adjustments for variability it is subjected to.
   - What else could be done with these?  
   - What other controls could be exploited? 
      - endogenous controls
      - sample controls

<br/>

### Before any modeling, fitting, etc. 

<p><p/>
* We need to examine the data to get a sense of what the problem is:
   - are there obvious quality issues?
   - are there less obvious quality issues?  
   - visually, how hard of a classification problem  do we have?
      - are cancer types distinct?
      - are there outlier points?  Is the outlier due to technical reasons?
 
<br/>

<!--
#### Word on modeling

* It is the error structure in the dataset that dictates the properties 
and validity of the computed statistics 

* Combining data points works well statistically if they are independent with
common mean and variance (+/-)

* statistical modeling is essentially the problem of finding                    contrasts among variables or transformations of variables such that
the contrasts are independent with common mean and sd.

-->



    
<!-- To run
## nohup Rscript -e "knitr::knit2html('index.Rmd')" > index.log  &
    
## Or 
## nohup Rscript -e "rmarkdown::render('index.Rmd')" > index.log  &
    
-->
