---
title: "Adela's cfMeDIP-seq Technology"
#subtitle:
author: "Francois Collin ([CV](https://francoiscollin-cv.netlify.app), [linkedin](https://www.linkedin.com/in/francoisz/))"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  #html_document:
  # for use of .tabset
  bookdown::html_document2:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 3
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../_bibFiles/_misc.bib, ../_bibFiles/_AI_ML.bib, ../_bibFiles/_RUV.bib, ../_bibFiles/_scRNAseq_norm.bib, ../_bibFiles/_ref_free_genom_inf.bib, ../_bibFiles/_healthy_aging.bib, ../../_bibFiles/_Normalization.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Freedman.bib, ../../_bibFiles/_Yu.bib,  ../../_bibFiles/_RMA.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

<p style="background-color: yellow; padding: 5px;">
  <span style="font-weight: bold;">⚠️ Work in Progress:</span> This content is incomplete.
</p>



```{css sidenote, echo = FALSE}

.main-container {
    margin-left: 250px;
}
.sidenote, .marginnote { 
  float: right;
  clear: right;
  margin-right: -40%;
  width: 37%;         # best between 50% and 60%
  margin-top: 0;
  margin-bottom: 0;
  font-size: 1.1rem;
  line-height: 1.3;
  vertical-align: baseline;
  position: relative;
  }
```


<style>
@import url('https://fonts.googleapis.com/css?family=Raleway');
@import url('https://fonts.googleapis.com/css?family=Oxygen');
@import url('https://fonts.googleapis.com/css?family=Raleway:bold');
@import url('https://fonts.googleapis.com/css?family=Oxygen:bold');

.main-container {
  max-width: 1400px !important;
}

body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1400px; }

caption {
  font-size: 20px;
  caption-side: top;
  text-indent: 30px;
  background-color: lightgrey;
  color: black;
  margin-top: 5px;
}

g-table-intro h4 {
  text-indent: 0px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```


```{r infex-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r index-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 


FN <- "index"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "index" #- replace by FLOWCELL???

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


```
<!-- ######################################################################## -->

<br/>


# About this page {-}

* In this page we collect comments and links to reports and studies
which speak to the development of Adela's cfMeDIP-seq Technology.
The source of all the information contained therein is public.


<br/>

# Reports   {-}


* [Adela vs GRAIL](AI/_Adela_vs_GRAIL_MCED_Comparison.html)

* [Evolution of Adela's Classifier]()
  



<br/>

# References {-}

<div id="refs"></div>

<br/>

<!--
# Appendix  {-}

<br/>
-->

```{r, echo=FALSE}
  knitr::knit_exit()
```


    
<!-- To run
## nohup Rscript -e "knitr::knit2html('index.Rmd')" > index.log  &
    
## Or 
## nohup Rscript -e "rmarkdown::render('index.Rmd')" > index.log  &
    

