---
title: "Adela's cfMeDIP-seq Technology"
#subtitle:
author: "Francois Collin ([CV](https://francoiscollin-cv.netlify.app), [linkedin](https://www.linkedin.com/in/francoisz/))"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  #html_document:
  # for use of .tabset
  rmarkdown::html_document:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 2
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [AI/_cfMeDIP-seq_references.bib, spike_ins/Ref/_spike_in.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

<p style="background-color: yellow; padding: 5px;">
  <span style="font-weight: bold;">⚠️ Work in Progress:</span> This content is incomplete.
</p>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```


```{r infex-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r index-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 


FN <- "index"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "index" #- replace by FLOWCELL???

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))

```
<!-- ######################################################################## -->

<br/>

# About this page {-}

* In this page we review the literature
which document the development of Adela's cfMeDIP-seq Technology.
The source of all the information contained therein is public.
The primary references are the following:^[
<span style="color:red">
[@Shen2018; @Wilson2022; @Hallaaho2022; @Zeng2023; @Stutheit-Zhao2024;
@Liu2025]
</span>
]

<div id="refs"></div>

<br/>

# AI Reports 

* AI reports are produced by claude.ai Opus 4.5 with 
naive prompts from Francois Collin.

* The reporting done is intended to provide the information
needed to understand the roots and development of 
Adela's cfMeDIP-seq Technology.


<br/>

## Adela vs GRAIL

* [slides](AI/_Adela_vs_GRAIL_MCED_Comparison_slides/_Adela_vs_GRAIL_MCED.html)

* [Rmarkdown](AI/_Adela_vs_GRAIL_MCED_Comparison.html)
   -  A fairly comprehensive, high level comparison of the two platforms.
   -  This compilation is for background purposes - see the documentation below 
for comments on technical aspects of some of the studies cited.

* There are several  other Dx commercial/venture companies going after methylation profiles
to detect cancer.  Can we have a table describing all companies targeting methylation.



<br/>

## cfMeDIP-seq: Protocol, Pipeline and Classifiers

* [slides](AI/_cfMeDIP-seq_Protocol_Pipeline_Classifiers_slides/_cfMeDIP_Evolution.html)

* [Rmarkdown](AI/_cfMeDIP-seq_Protocol_Pipeline_Classifiers.html) 
   - The evolution of Adela's cfMeDIP-seq Protocol, Pipeline and Classifiers is 
traced.
   - The report is a fairly comprehensive, high level description of the assay
and algorithm along with a description of enhancements that followed.

<br/>

### Supplementary Resources {.tabset}

#### Home {-}

<br/>

#### Code Repositories {-}

| Resource | URL | Description |
|----------|-----|-------------|
| MEDIPIPE | https://github.com/pughlab/MEDIPIPE | Main analysis pipeline |
| cfMeDIP-seq analysis | https://github.com/pughlab/cfMeDIP-seq-analysis-pipeline | Post-processing pipeline |
| Halla-aho scripts | https://github.com/hallav/cfMeDIP-seq | Classifier comparison code |
| Shen et al. ML models | http://doi.org/10.5281/zenodo.1242697 | Original classifier code |
| Protocol tables | https://github.com/bratmanlab/cfMeDIP_Protocol | Processed data tables |

<br/>

####  Public Data Repositories {-}

| Dataset | Repository | Accession | Access |
|---------|------------|-----------|--------|
| HCT116 cell line | GEO | GSE79838 | Public |
| Brain tumors | Zenodo | (see Nassiri et al.) | Public |
| Healthy controls | EGA | EGAD50000000652 | Controlled |
| INSPIRE study | EGA | EGAD00001011312 | Controlled |

<br/>

## spike-in control normalization 


* [slides](spike_ins/_Wilson_2022_Spike_in_Controls_Technical_Review_slides/_Wilson2022_SpikeIns.html)
* [Rmarkdown](spike_ins/_Wilson_2022_Spike_in_Controls_Technical_Review.html)
   - contents of the paper are detailed.
   - [Figures](spike_ins/_Wilson_2022_Figures_Captions.html)

<br/>

### Additional Resources {.tabset}

#### Home {-}

<br/>

#### Additional markdowns {-} 

* [framework](spike_ins/_spike_in_normalization_cfMeDIP_report.html)
* [biological validation](spike_ins/_Wilson_2022_Critical_Analysis_Biological_Validation.html)

* [NGS_Normalization_Methods_Review](spike_ins/NGS_Normalization_Methods_Review.html)


<br/>

####   Published Data {-}

* **GEO Accession GSE166259:**
   - Raw HCT116 cell line data (public)
   - Processed AML patient data (public)

* **EGA Accession EGAS00001005069:**
   - Raw AML patient data (controlled access)
   - Available via University Health Network Genomics Data Access Committee

* **Zenodo Deposits:**
   - Code for reproducing all analyses: https://doi.org/10.5281/zenodo.4683791
   - spiky package: https://doi.org/10.18129/B9.bioc.spiky

<br/>
<br/>

## Fragment length distribution

* Stutheit-Zhao (2024) [@Stutheit-Zhao2024] show that the 
distribution of fragment lengths adds information to sequence
data when trying to discriminate  between cancer and non-cance samples.


**TO BE COMPLETED** - put summary here.

<br/>
<br/>

# Re-analysis 

* Two areas in the developmental path that are inviting to further
study: 

   - Spike-in Normalization and 
   - Cancer Model Selection

<br/>

## Spike-in Normalization {.tabset}

### Home {-}

<br/>

<!--
* [Wilson2022_aws](AWS/Wilson2022_aws) - setup docs
-->

### EC2 Install  {-}

* [EC2_SETUP_GUIDE](AWS/Wilson2022_aws/EC2_SETUP_GUIDE.html)

* [EC2_TROUBLESHOOTING_GUIDE](AWS/Wilson2022_aws/EC2_TROUBLESHOOTING_GUIDE.html)

* [master_setup](AWS/Wilson2022_aws/master_setup.html)

* [QUICK_REFERENCE](AWS/Wilson2022_aws/QUICK_REFERENCE.html)

* [WILSON_PIPELINE](AWS/Wilson2022_aws/WILSON_PIPELINE.html)

<br/>
<br/>

 
### Data and Code {-}

* [Data Download and Organization Guide](spike_ins/_Wilson2022_Data_Download_Guide.html)

<br/>


### Re-analysis {-}

* [_cfmedip_normalization_presentation](spike_ins/new_slides/_cfmedip_normalization_presentation.html)


<br/>

## Cancer Model Selection

*  Examine more  model structures - enet, rf, pam, svm, gbm, knn, xgb
   - in addition to the error rate, we need to examine
where the errors occur (samples) and how (miss by a lot or little)
   - compare validation and test error rates.

<br/>
<!--

# More recent publications {.tabset}

## Home {-}

<br/>

## Stutheit-Zhao et al. (2024) {-}

* claude.ai sonnet


<br/>

## Liu et al. (2025)  {-}

* See **Clinical Validation Studies for molDx Products**

-->

<br/>

# Extras {.tabset}

## Home {-}

<br/>

## DMA {-}
 

* [Differential Methylation Analysis Methodology](AI/_Differential_Methylation_Analysis_Report.html)

<p/>
* Exhaustive survey of DMA methods used with each  assay type. 
   - Not specific to Adela
* Describe the analysis data in each step
* Survey of benchmarking studies and tools

<br/>

## Enrichment Based DNA {-}


* [Enrichment-Based Differential Methylation Analysis](AI/_Enrichment_Methods_PartI_Section1.html)

<p/>
* For enrichment-Based assays, a look at Bioinformatics Pipelines and 
Statistical Methods for Differential Methylation Analysis: 
   - Common Bioinformatics Pipeline:FASTQ->BAN[ ](AI/_Enrichment_Methods_PartI_Section1.html#common-bioinformatics-pipeline-fastq-to-bam)
   - DMA Methods for Enrichment based assays[ ](AI/_Enrichment_Methods_PartI_Section1.html#dma-methods)
      - many of the methods produce estimates of `absolute beta`, a measure of
methylation level.  These would then be further analyzed to
get DMA results.

<br/>
 

## Variability in  cfMeDIP-seq Data {-}

<p/>
* The key components of the assay - filler DNA and antibody instance - 
are associated with a considerable amount of variability.
   - [docname](./docname.html) provides a detailed look at
the Methylated DNA Immunoprecipitation Stage of the protocol.
   - questions to address:
      - how does the filler DNA operate?  where does the variability
come from?
      - how does anti-5-methylcytosine (5mC) antibody work?
where does the variablity come from?
      - can variability be attributed to its proper source -
anti-5-methylcytosine (5mC) antibody or filler DNA?

<div style="color:green">

*  Rerun the analyses which assess the performance of the
spike-in controls.
   -  Wilson et al. (2022) [@Wilson2022]

</div>

<!--
## Companies targeting methylation
-->

```{r,  child="methylation_cancer_dx_landscape.md"}
```





<br/>

# References 

<div id="refs"></div>

<br/>


```{r, echo=FALSE}
  knitr::knit_exit()
```

############################################################
# ARCHIVED CODE BELOW
############################################################

#bibliography_cfMeDIP1: [AI/_cfMeDIP-seq_references_1.bib]
#bibliography_cfMeDIP2: [AI/_cfMeDIP-seq_references_2.bib]
#bibliography_cfMeDIP3: [AI/_cfMeDIP-seq_references_2.bib]
#bibliography_spikeins: [spike_ins/Ref/_spike_in.bib]


\nocite{Burgener2021,DePascali2024,Hallaaho2022,Liu2025,Nassiri2020,Nuzzo2019,Nuzzo2020,Shen2018,Shen2019,Stutheit-Zhao2024,Wilson2022,Zeng2023}

<div id="refs"></div>

### Norm Background {-}

* Normalization is one of the steps in the analysis of sample cfMeDIP data
towards cancer detection scores.
* Two models are used, each has the binned coverage as input
   - first model predicts the probability of cancer
   - second model predicts the cancer type
* a 27 fragment spike-in normalization is suggested to adjust coverage levels
   - a scaling factor adjusted for 3 covariates is suggested
   - 4 parameters computed from 27 points which may or may not represent
the sample distribution

* In the task of normalization, we need to counterweigh
the desire to have unbiased estimates with the need to have
low variability estimates.
   - the data need to be examined to see we are on the continuum.
   - a 27 fragment spike-in normalization is going to be on the noisy side

* Alternatively, identify 1000 stably expressed bins and use some one-poarameter
summary of these as shift.

<br/>

# Appendix: Full list of reports

```{r , child=c('_index.Rmd')}
```


## Full List of AI Reports


```{r , child=c('_index_AI.Rmd')}
```

<br/>

### Study by Study Execution Outline  {-}

#### **Shen et al. (2018) [@Shen2018]** {-}

* Was the Differentially Methylated Regions (DMRs) status between cancer and controls
used for selecting features?

* CV is to determine the hyper-parameters

```
Training:

* 10-fold cross-validation for hyper-parameter tuning
* Nested cross-validation for performance estimation
* Class imbalance handled by:
   - Weighted loss function
   - Or downsampling of majority class

```
 
* class imbalance is not an issue when comparing algorithms based on the same sample set.

<p><p/>
* auROC is not a good summary
   - spec at fixed sens would be better
   - but both would order algorithms similarly.



<br/>

### Controls, Controls, Controls 

<p><p/>
* Spike-In Controls - Wilson et al. (2022) [@Wilson2022]
   - How are the Synthetic Spike-In Controls used to reduce unwanted variability?


   - Spike-ins may not be affected by some important sources of variability,
but they can only contribute adjustments for variability it is subjected to.
   - What else could be done with these?  
   - What other controls could be exploited? 
      - endogenous controls
      - sample controls

<br/>

### Before any modeling, fitting, etc. 

<p><p/>
* We need to examine the data to get a sense of what the problem is:
   - are there obvious quality issues?
   - are there less obvious quality issues?  
   - visually, how hard of a classification problem  do we have?
      - are cancer types distinct?
      - are there outlier points?  Is the outlier due to technical reasons?
 
<br/>

<!--
#### Word on modeling

* It is the error structure in the dataset that dictates the properties 
and validity of the computed statistics 

* Combining data points works well statistically if they are independent with
common mean and variance (+/-)

* statistical modeling is essentially the problem of finding                    contrasts among variables or transformations of variables such that
the contrasts are independent with common mean and sd.

-->


## Full List of AI Reports


```{r , child=c('_index_AI.Rmd')}
```

<br/>

### Study by Study Execution Outline  {-}

#### **Shen et al. (2018) [@Shen2018]** {-}

* Was the Differentially Methylated Regions (DMRs) status between cancer and controls
used for selecting features?

* CV is to determine the hyper-parameters

```
Training:

* 10-fold cross-validation for hyper-parameter tuning
* Nested cross-validation for performance estimation
* Class imbalance handled by:
   - Weighted loss function
   - Or downsampling of majority class

```
 
* class imbalance is not an issue when comparing algorithms based on the same sample set.

<p><p/>
* auROC is not a good summary
   - spec at fixed sens would be better
   - but both would order algorithms similarly.



<br/>

### Controls, Controls, Controls 

<p><p/>
* Spike-In Controls - Wilson et al. (2022) [@Wilson2022]
   - How are the Synthetic Spike-In Controls used to reduce unwanted variability?


   - Spike-ins may not be affected by some important sources of variability,
but they can only contribute adjustments for variability it is subjected to.
   - What else could be done with these?  
   - What other controls could be exploited? 
      - endogenous controls
      - sample controls

<br/>

### Before any modeling, fitting, etc. 

<p><p/>
* We need to examine the data to get a sense of what the problem is:
   - are there obvious quality issues?
   - are there less obvious quality issues?  
   - visually, how hard of a classification problem  do we have?
      - are cancer types distinct?
      - are there outlier points?  Is the outlier due to technical reasons?
 
<br/>

<!--
#### Word on modeling

* It is the error structure in the dataset that dictates the properties 
and validity of the computed statistics 

* Combining data points works well statistically if they are independent with
common mean and variance (+/-)

* statistical modeling is essentially the problem of finding                    contrasts among variables or transformations of variables such that
the contrasts are independent with common mean and sd.

-->



    
<!-- To run
## nohup Rscript -e "knitr::knit2html('index.Rmd')" > index.log  &
    
## Or 
## nohup Rscript -e "rmarkdown::render('index.Rmd')" > index.log  &
    
-->
