---
title: "Adela's cfMeDIP-seq Technology"
#subtitle:
author: "Francois Collin ([CV](https://francoiscollin-cv.netlify.app), [linkedin](https://www.linkedin.com/in/francoisz/))"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  #html_document:
  # for use of .tabset
  rmarkdown::html_document:
    code_folding: hide
    code_download: true
    toc: false
    toc_depth: 3
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../_bibFiles/_cfMeDIP-seq.bib, ../_bibFiles/_misc.bib, ../_bibFiles/_AI_ML.bib, ../_bibFiles/_RUV.bib, ../_bibFiles/_scRNAseq_norm.bib, ../_bibFiles/_ref_free_genom_inf.bib, ../_bibFiles/_healthy_aging.bib, ../../_bibFiles/_Normalization.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Freedman.bib, ../../_bibFiles/_Yu.bib,  ../../_bibFiles/_RMA.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

<p style="background-color: yellow; padding: 5px;">
  <span style="font-weight: bold;">⚠️ Work in Progress:</span> This content is incomplete.
</p>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```


```{r infex-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r index-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 


FN <- "index"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "index" #- replace by FLOWCELL???

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


```
<!-- ######################################################################## -->

<br/>

<!--
* Donoho on Frictionless Reproducibility
   - Donoho on the Modern History of Stats
-->

### About this page {-}

* In this page we collect comments and links to reports and studies
which speak to the development of Adela's cfMeDIP-seq Technology.
The source of all the information contained therein is public.


<br/>

### Reports {-}

<p><p/>
* [Adela vs GRAIL](AI/_Adela_vs_GRAIL_MCED_Comparison.html)
   - A fairly comprehensive, high level comparison of the two platforms.
   - This compilation is for background purpopses - see the documentation below for comments on
technical aspects of some of the studies cited.

<br/>

<p><p/>
* [Evolution of Adela's cfMeDIP-seq Protocol, Pipeline and Classifiers](AI/_cfMeDIP-seq_Protocol_Pipeline_Classifiers.html)
   - A fairly comprehensive, high level description of the assay and algorithm along with
a description of enhancements that followed.
   - Comments will be made when the primary references are described below.


<br/>

<p><p/>
* [Comments on Evolution of Adela's cfMeDIP-seq Protocol, Pipeline and Classifiers](AI/_cfMeDIP-seq_Protocol_Pipeline_Classi_comments.html) -
   - Some comments have to do with potential leakage; only by going through the code
will we ever be able to figure this out.
   - Other comments have to do with evidence gathering.


<br/>

### Study by Study Execution Outline  {-}

#### **Shen et al. (2018) [@Shen:2018aa]** {-}

* Was the Differentially Methylated Regions (DMRs) status between cancer and controls
used for selecting features?

* CV is to determine the hyper-parameters

```
Training:

* 10-fold cross-validation for hyper-parameter tuning
* Nested cross-validation for performance estimation
* Class imbalance handled by:
   - Weighted loss function
   - Or downsampling of majority class

```
 
* class imbalance is not an issue when comparing algorithms based on the same sample set.

<p><p/>
* auROC is not a good summary
   - spec at fixed sens would be better
   - but both would order algorithms similarly.



<br/>

### Controls, Controls, Controls 

<p><p/>
* Spike-In Controls - Wilson et al. (2022) [@Wilson:2022aa]
   - How are the Synthetic Spike-In Controls used to reduce unwanted variability?


   - Spike-ins may not be affected by some important sources of variability,
but they can only contribute adjustments for variability it is subjected to.
   - What else could be done with these?  
   - What other controls could be exploited? 
      - endogenous controls
      - sample controls

<br/>

### Before any modeling, fitting, etc. 

<p><p/>
* We need to examine the data:
   - are there obvious quality issues?
   - are there less obvious quality issues?  
   - visually, how hard of a classification problem  do we have?
      - are cancer types distinct?
      - are there outlier points?  Is the outlier due to technical reasons?
 
<br/>

### elastic net Vs RF 



<br/>

# References {-}

<div id="refs"></div>

<br/>

<!--
# Appendix  {-}

<br/>
-->

```{r, echo=FALSE}
  knitr::knit_exit()
```


    
<!-- To run
## nohup Rscript -e "knitr::knit2html('index.Rmd')" > index.log  &
    
## Or 
## nohup Rscript -e "rmarkdown::render('index.Rmd')" > index.log  &
    
-->
