---
title: "cfMeDIP-seq Evolution"
subtitle: "Protocol, Pipeline, and Classifier Methodologies"
author: "Technical Evolution Timeline"
date: "January 2026"
output:
  revealjs::revealjs_presentation:
    theme: sky
    highlight: pygments
    center: true
    transition: slide
    css: custom_cfmedip.css
    reveal_options:
      slideNumber: true
      previewLinks: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Executive Summary {.smaller}

This presentation traces the evolution of **cell-free methylated DNA immunoprecipitation sequencing (cfMeDIP-seq)** methodology:

:::: {.columns}
::: {.column width="50%"}
**Protocol Development**
- Nature Protocols 2019
- Variability mitigation
- Sample format optimization
:::

::: {.column width="50%"}
**Computational Evolution**
- MEDIPIPE pipeline
- Classifier methodologies
- Machine learning advances
:::
::::

<br>

> From foundational academic protocol to clinical-grade assay platform

# 1. Protocol Foundation

## cfMeDIP-seq Protocol Overview

**Published: Nature Protocols 2019** (Shen et al.)

**Key Innovation:** Adaptation of MeDIP-seq for **low-input cell-free DNA**

- Input requirement: **1-10 ng** cfDNA
- Timeline: **3-4 days** (standard molecular biology lab)
- Core innovation: **Filler DNA** as carrier molecule

<br>

**Filler DNA Function:**
- Improves antibody-mediated capture efficiency
- Enhances specificity for methylated fragments
- Enables ultra-low input processing

## Protocol Stages

**Four Major Stages:**

1. **Library Preparation** (Day 1) - Steps 1-3
   - End-repair & A-tailing
   - Adapter ligation
   - SPRI bead purification

2. **Immunoprecipitation** (Day 2) - Steps 4-10 ⚠️
   - Add filler DNA & spike-ins
   - Denature DNA (95°C)
   - Anti-5mC antibody IP (overnight)
   - Bead capture & wash
   - Elution & purification

## Protocol Stages (continued)

3. **Library Amplification** (Day 3) - Steps 11-14
   - qPCR quality control
   - PCR amplification
   - Size selection & QC

4. **High-Throughput Sequencing** (Day 4+) - Step 15
   - Paired-end sequencing
   - Data generation

<br>

⚠️ **Critical Stage:** Immunoprecipitation (Stage 2) is the core of the assay and primary source of technical variability

## Sources of Variability

**HIGH Variability Sources:**

- **Filler DNA composition** (methylated vs unmethylated)
- **Antibody lot variation** (batch-to-batch differences)
- **Immunoprecipitation efficiency** (overnight incubation)

**MODERATE Variability:**

- Adapter ligation efficiency
- PCR amplification (GC bias, cycle number)
- Sequencing depth variations

**LOW Variability:**

- End-repair & A-tailing
- SPRI bead purification
- Size selection

# 2. Variability Mitigation

## Systematic vs Random Variability

:::: {.columns}
::: {.column width="50%"}
**Systematic (Correctable)**

✓ Batch effects  
✓ GC bias  
✓ Mappability bias  
✓ Fragment length bias  

*Solution: Spike-in normalization*
:::

::: {.column width="50%"}
**Random (Design-dependent)**

⚠ Antibody lot variation  
⚠ IP efficiency fluctuation  
⚠ Library prep stochasticity  

*Solution: Experimental controls*
:::
::::

<br>

**Key Improvement: Wilson et al. 2022**
- Spike-in normalization methodology
- Batch effect reduction: 5% → <1% variance
- Published in *Cell Reports Methods*

## Protocol Improvements Timeline {.smaller}

| Year | Development | Impact |
|------|-------------|--------|
| **2018** | Foundational methodology (Shen et al., *Nature*) | Established cfMeDIP-seq feasibility |
| **2019** | Standardized protocol (Shen et al., *Nat Protoc*) | Reproducible 3-4 day workflow |
| **2020** | Disease-specific applications (Nassiri, Nuzzo) | Brain tumors, RCC detection |
| **2022** | Spike-in normalization (Wilson et al.) | Batch variance: 5% → <1% |
| **2023** | MEDIPIPE automation (Zeng et al.) | End-to-end computational pipeline |
| **2024** | Clinical validation (multiple) | MRD, MCED applications |

## Critical Batch Boundaries

**Randomization is Essential at These Points:**

1. **cfDNA extraction from plasma**
   - Different extraction dates
   - Different technicians

2. **Library preparation batches**
   - Filler DNA aliquots
   - Adapter ligations

3. **Immunoprecipitation batches**
   - Antibody lots
   - Different IP dates

4. **PCR amplification**
   - Thermocycler variation
   - Primer lots

5. **Sequencing runs**
   - Different flow cells
   - Different instruments

# 3. MEDIPIPE Pipeline

## Pipeline Overview

**MEDIPIPE: Automated cfMeDIP-seq Analysis**

Published: *Bioinformatics* 2023 (Zeng et al.)

**Three Core Modules:**

1. **Quality Control** - FastQC, alignment metrics, enrichment validation
2. **Methylation Quantification** - Genome-wide binning, count matrices
3. **Sample Aggregation** - Multi-sample normalization, batch correction

<br>

> "Automated and comprehensive pipeline for cfMeDIP-seq data quality control and analysis"

## Pipeline Architecture

**Input:** Paired-end FASTQ files

**Processing Steps:**

- Adapter trimming (Trim Galore!)
- Alignment to reference genome (Bowtie2)
- Duplicate marking (Picard)
- Quality metrics computation
- Coverage analysis across genomic features
- Generation of normalized count matrices

**Output:**

- Sample-level QC reports
- Genome-wide methylation profiles
- Statistical input matrices for classification

## Key Pipeline Features {.smaller}

**1. Comprehensive QC Metrics:**
- Fragment size distribution
- Enrichment over CpG islands
- Signal-to-noise ratios
- Library complexity

**2. Flexible Binning Strategies:**
- Fixed-width genomic windows (300 bp, 500 bp, 1 kb)
- Feature-based regions (CpG islands, shores, shelves)
- FANTOM5 enhancer regions
- Custom region sets

**3. Normalization Methods:**
- Read depth normalization (CPM, RPM)
- GC content correction
- Mappability adjustment
- Spike-in normalization integration

# 4. Classifier Evolution

## Foundational Classifier (2018)

**Shen et al. 2018 (*Nature*) - Original Methodology**

**Feature Engineering:**
- 300 bp genomic windows
- Mapped to: CpG islands, shores, shelves, FANTOM5 enhancers
- Top 1,000 most variable fragments

**Classification Algorithms:**
- **GLMnet** (elastic net regularization) ← Primary
- **Random Forest**

**Performance:**
- Successfully distinguished cancer from healthy controls
- Enabled multi-cancer type classification
- Tissue-of-origin prediction capability

## Feature Selection Evolution {.smaller}

**Multiple Approaches Explored:**

| Method | Purpose | Performance |
|--------|---------|-------------|
| t-tests | Statistical significance testing | Moderate |
| Fisher's exact test | Feature selection | **Robust** ✓ |
| PCA | Dimensionality reduction | Standard |
| Iterative supervised PCA (ISPCA) | Feature generation | **Promising** ✓ |
| GLMnet | L1/L2 regularization | **Robust** ✓ |
| Logistic regression | Sparsity-promoting priors | **Simple & effective** ✓ |

**Best Practices (Halla-aho & Lähdesmäki 2022):**
- Fisher's exact test for feature selection
- ISPCA for feature generation  
- Logistic regression with hyper/hypomethylated region counts

## Application-Specific Classifiers {.smaller}

**Renal Cell Carcinoma (Nuzzo et al. 2019/2020):**
- Top 300 differentially methylated regions (DMRs)
- Methylation score assignment
- Validated in plasma and urine samples

**Breast Cancer / BRCA1/2 (De Pascali et al. 2024):**
- GLMnet and Random Forest models
- BRCA1/2 mutation detection capability
- Breast cancer diagnosis from cfDNA

**Multi-Cancer Platform (Adela):**
- Same wet-lab assay across applications
- Classifier-specific development per application
- Enables rapid scaling to new cancer types

> "Platform efficiency: same assay, different classifiers"

## Classifier Methodology Comparison {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Early Approaches**
- Manual feature selection
- Limited genomic regions
- Single cancer focus
- Cross-validation within single cohorts
:::

::: {.column width="50%"}
**Modern Approaches**
- Automated feature engineering
- Genome-wide coverage
- Multi-cancer capability
- Independent validation sets
- Clinical utility validation
:::
::::

<br>

**Key Evolution:** From technical feasibility → clinical validation

# 5. Platform Comparisons

## cfMeDIP-seq vs Bisulfite Sequencing {.smaller}

| Feature | cfMeDIP-seq | Bisulfite Sequencing |
|---------|-------------|----------------------|
| **DNA Modification** | None (preservation) | Chemical conversion |
| **DNA Degradation** | Minimal | Substantial (~95% loss) |
| **Input Requirement** | 1-10 ng | 10-100 ng |
| **Resolution** | Regional (~300 bp) | Single-base (CpG-level) |
| **Coverage** | Genome-wide enrichment | Targeted or whole-genome |
| **Cost per Sample** | Lower | Higher (WGBS) |
| **Library Complexity** | High | Reduced (4 nucleotides → 3) |
| **Ideal Application** | Low cfDNA, genome-wide | High cfDNA, specific CpGs |

## cfMeDIP-seq vs Targeted Methods

**Targeted Methylation Arrays:**
- Fixed CpG coverage (e.g., 850K array)
- No discovery capability
- Lower cost per sample
- High throughput

**cfMeDIP-seq Advantages:**
- Flexible genomic interrogation
- Novel biomarker discovery
- Captures regulatory elements
- Adapts to new genomic annotations

**Use Case Optimization:**
- Arrays: Known biomarker validation, large cohorts
- cfMeDIP-seq: Discovery research, low-input samples, MRD

# 6. Sample Processing

## Physical Sample Format Transitions

**Plasma → cfDNA:**
- Blood draw (10 mL)
- Plasma separation (centrifugation)
- cfDNA extraction (column-based)
- Quantification (fluorometry)
- **Typical yield:** 10-50 ng/mL plasma

**cfDNA → Library:**
- End-repair & A-tailing (tube)
- Adapter ligation (tube)
- **+ Filler DNA addition** ← Key innovation
- Purification (magnetic beads)

## Sample Format Transitions (continued)

**Library → Enriched Library:**
- Denaturation (thermocycler)
- Immunoprecipitation (overnight, tube)
- Magnetic bead capture
- Washes (multiple)
- Elution (purified methylated DNA)

**Enriched Library → Sequencing Library:**
- qPCR quality assessment
- PCR amplification (8-12 cycles)
- Size selection (SPRI beads)
- Final quantification

**Sequencing Library → Data:**
- Paired-end sequencing (Illumina)
- **Typical output:** 20-40 million read pairs

## Typical Batch Structure

**Single Processing Batch:**

- 24-96 samples typical
- Include controls:
  - Positive controls (methylated gDNA)
  - Negative controls (unmethylated λDNA)
  - Technical replicates (10-20%)
  - Spike-in standards

**Randomization Strategy:**
- Stratify by: cancer type, stage, treatment status
- Distribute across: extraction dates, IP batches, sequencing runs
- Balance: case/control ratio per batch

# 7. Best Practices

## Protocol Best Practices {.smaller}

**Sample Handling:**
1. Process plasma within 2 hours of blood draw
2. Double-spin plasma to remove cellular debris
3. Store cfDNA at -80°C (avoid freeze-thaw)
4. Quantify DNA accurately (Qubit fluorometry)

**Library Preparation:**
5. Use same filler DNA lot within study
6. Perform IP with consistent antibody lots
7. Include spike-ins for normalization
8. Monitor enrichment with qPCR

**Quality Control:**
9. Fragment size: 150-200 bp peak
10. Library complexity: >80% unique reads
11. Enrichment: >5-fold over input
12. Sequencing depth: ≥20M paired reads

## Computational Best Practices

**Data Processing:**
- Use MEDIPIPE or equivalent automated pipeline
- Apply adapter trimming (quality score >20)
- Align with Bowtie2 (default parameters)
- Remove PCR duplicates

**Feature Generation:**
- Bin reads into 300-500 bp windows
- Focus on CpG-rich regions
- Include enhancer elements (FANTOM5)
- Normalize for GC content and mappability

**Statistical Analysis:**
- Apply spike-in normalization
- Use robust feature selection (Fisher's exact, ISPCA)
- Validate with independent test sets
- Report performance by cancer stage

## Future Research Priorities

**Technical Development:**

1. **Single-cell cfMeDIP-seq**
   - Ultra-low input optimization
   - Single-molecule resolution

2. **Improved spike-in strategies**
   - Synthetic methylated standards
   - Quantitative calibration curves

3. **Automation & standardization**
   - Robotic liquid handling
   - Pre-aliquoted reagent kits

## Future Priorities (continued)

**Clinical Translation:**

4. **Prospective trial validation**
   - Multi-center studies
   - Real-world performance data

5. **Reimbursement pathway**
   - Health economics studies
   - Clinical utility demonstration

6. **Regulatory approval**
   - FDA clearance pathways
   - Clinical trial designs

# 8. Key Publications

## Foundational Papers {.smaller}

**Protocol & Methodology:**

1. **Shen et al. 2018** - *Nature* 563:579-583
   - "Sensitive tumour detection and classification using plasma cell-free DNA methylomes"
   - Established foundational methodology

2. **Shen et al. 2019** - *Nat Protoc* 14:2749-2780
   - "Preparation of cfMeDIP-seq libraries for methylome profiling"
   - Standardized reproducible protocol

3. **Wilson et al. 2022** - *Cell Rep Methods*
   - Spike-in normalization methodology
   - Batch effect reduction from 5% to <1%

## Application Papers {.smaller}

**Disease-Specific Validation:**

4. **Nassiri et al. 2020** - *Nat Med*
   - "Detection and discrimination of intracranial tumors"
   - Brain tumor methylome profiling

5. **Nuzzo et al. 2020** - *Nat Med* 26:1041-1043
   - "Detection of renal cell carcinoma using plasma and urine cfDNA methylomes"
   - RCC detection from liquid biopsies

6. **De Pascali et al. 2024** - *J Transl Med* 22:934
   - "Differential methylation for breast cancer diagnosis and BRCA1/2 detection"
   - Breast cancer application

## Computational Tools {.smaller}

**Bioinformatics Development:**

7. **Zeng et al. 2023** - *Bioinformatics* 39:btad423
   - "MEDIPIPE: automated and comprehensive pipeline"
   - End-to-end analysis automation

8. **Halla-aho & Lähdesmäki 2022** - *BMC Bioinformatics* 23:138
   - "Probabilistic modeling methods for cfDNA methylation"
   - Comprehensive classifier comparison

**Available Resources:**
- GitHub: `pughlab/MEDIPIPE`
- GitHub: `pughlab/cfMeDIP-seq-analysis-pipeline`
- GitHub: `hallav/cfMeDIP-seq`

# 9. Public Data Resources

## Available Datasets {.smaller}

**Public Repositories:**

| Dataset | Accession | Description |
|---------|-----------|-------------|
| Cell line data | GEO: GSE79838 | Shen et al. 2018 validation |
| Healthy controls | EGA: EGAD50000000652 | 23 plasma cfDNA samples |
| INSPIRE study | EGA: EGAD00001011312 | Multi-timepoint immunotherapy |
| Brain tumor demo | Zenodo | 163-sample demonstration set |

**Machine Learning Models:**

- Zenodo: `10.5281/zenodo.1242697` (Shen et al. 2018 models)
- Zenodo: `10.5281/zenodo.1490920` (Intermediate objects)

**Note:** Most patient data requires controlled access (institutional ethics compliance)

# 10. Evolution Summary

## Timeline of Innovation

```{r evolution-timeline, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6}
library(ggplot2)
library(dplyr)

timeline_data <- data.frame(
  Year = c(2018, 2019, 2020, 2022, 2023, 2024),
  Event = c(
    "Foundation\n(Nature)",
    "Protocol\n(Nat Protoc)",
    "Applications\n(Brain/RCC)",
    "Normalization\n(Spike-in)",
    "Automation\n(MEDIPIPE)",
    "Clinical\n(MRD/MCED)"
  ),
  Category = c("Method", "Protocol", "Application", "Improvement", "Software", "Clinical"),
  Impact = c(10, 8, 7, 9, 8, 10)
)

ggplot(timeline_data, aes(x = Year, y = Impact, color = Category)) +
  geom_line(size = 1.5, alpha = 0.5) +
  geom_point(size = 8, alpha = 0.8) +
  geom_text(aes(label = Event), vjust = -1.5, size = 3.5, fontface = "bold") +
  scale_y_continuous(limits = c(0, 12), breaks = seq(0, 10, 2)) +
  scale_x_continuous(breaks = seq(2018, 2024, 1)) +
  labs(
    title = "cfMeDIP-seq Evolution Timeline",
    subtitle = "From Academic Method to Clinical Platform",
    x = "Year",
    y = "Impact Level",
    color = "Development Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

## Key Takeaways {.smaller}

**Technical Evolution:**
1. Low-input cfDNA methodology (1-10 ng) enables liquid biopsy
2. Spike-in normalization reduces batch effects from 5% to <1%
3. MEDIPIPE automation standardizes analysis across labs
4. Multiple classifier approaches validated for different applications

**Platform Advantages:**
- Preserves DNA integrity (no bisulfite)
- Genome-wide discovery capability
- Flexible classifier development
- Single assay, multiple applications

**Clinical Translation:**
- MRD: Validated in head & neck, lung cancer
- MCED: 12 cancers in development, 50+ planned
- Performance: 92-97% sensitivity across stages

## Future Outlook

**Near-Term (1-2 years):**
- Complete CAMPERR study validation
- Expand MRD to additional cancer types
- Regulatory pathway progress

**Mid-Term (3-5 years):**
- FDA approval for MRD applications
- MCED clinical utility demonstration
- Integration into standard care pathways

**Long-Term (5+ years):**
- Population screening programs
- Multi-omics integration (cfDNA + protein)
- Personalized treatment monitoring

> From research tool to standard of care

# Conclusions

## Summary

cfMeDIP-seq has evolved from an academic methylation profiling method to a clinically-validated platform for cancer detection and monitoring.

**Key Milestones:**
- ✓ Foundational methodology (2018)
- ✓ Standardized protocol (2019)
- ✓ Variability mitigation (2022)
- ✓ Computational automation (2023)
- ✓ Clinical validation (2024)

**Enabling Factors:**
- Low DNA input requirements
- Genome-wide coverage
- Platform flexibility
- Robust normalization strategies

## Questions?

<br><br>

### Additional Resources

**Comprehensive Documentation:**
- Protocol: Shen et al., *Nat Protoc* 2019
- Pipeline: `github.com/pughlab/MEDIPIPE`
- Reviews: Multiple disease-specific validations

**Data Access:**
- Public: GEO, Zenodo
- Controlled: EGA (patient data)

<br>

*Presentation based on comprehensive technical review (January 2026)*

---

**Disclaimer:** Information for research and educational purposes. Clinical decisions should be made in consultation with healthcare providers.


<!--
nohup Rscript -e "rmarkdown::render('_cfMeDIP_Evolution.Rmd')" > _cfMeDIP_Evolution.log  &
     
-->
